---
title: "A FAFB left-right bridging registration from matched neuron tracts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A FAFB left-right bridging registration from matched neuron tracts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
# Deformetrica (>= 4.3) + a GPU are needed for the fit/shoot chunks (eval=FALSE).
```

## Goal

Build a **left-right (L-R) bridging registration** of the fly brain: a
diffeomorphism that maps one hemisphere onto the mirror image of the other, so a
neuron on the left can be compared directly with its cognate on the right. This
is the modern, package-native re-implementation of the recipe in
[flyconnectome/deformetricaLR](https://github.com/flyconnectome/deformetricaLR),
which fit one diffeomorphism to a **set of matched cognate neuron tracts** (each a
`NonOrientedPolyLine`), optionally anchored by neuropil **landmarks**.

The pipeline is three stages, following that repo and the package's `deform/`
scripts:

1. **Mirror** the right-side neurons to the left (a whole-brain flip).
2. **Affine** initialise (Morpho) so the flipped tracts roughly overlie their
   left cognates.
3. **Deformable** multi-object Deformetrica fit over all matched pairs at once,
   with a neuropil landmark object as a global regulariser.

```{r setup}
library(deformetricar)
library(nat)
library(nat.flybrains)         # FCWB template + mirroring
library(nat.templatebrains)    # mirror_brain
library(Morpho)
```

## 1. Matched cognate neurons, mirrored to one side

`nat::Cell07PNs` are 40 projection neurons in FCWB space with a `Glomerulus`
label — an ideal small, self-contained set of *typed* neurons whose left/right
cognates are the same glomerulus. We take the right cell of each glomerulus as
the moving template and its left cognate as the fixed target, and flip the right
cells to the left hemisphere first.

```{r matched}
pn <- Cell07PNs
right <- subset(pn, Side == "R")
left  <- subset(pn, Side == "L")
gloms <- intersect(right[, "Glomerulus"], left[, "Glomerulus"])

## flip the right neurons across the midline into left space (the L-R prior)
right_flipped <- nat.templatebrains::mirror_brain(right, brain = nat.flybrains::FCWB)

## match by glomerulus -> named lists of cognate pairs
src <- setNames(lapply(gloms, function(g) right_flipped[[which(right[, "Glomerulus"] == g)[1]]]), gloms)
tgt <- setNames(lapply(gloms, function(g) left[[which(left[, "Glomerulus"] == g)[1]]]), gloms)
```

## 2. Affine initialisation (Morpho)

Bring the flipped-right tracts onto the left cognates with a single affine before
the diffeomorphism (the flip is only approximately symmetric).

```{r affine}
P_src <- do.call(rbind, lapply(src, nat::xyzmatrix))
P_tgt <- do.call(rbind, lapply(tgt, nat::xyzmatrix))
# a coarse affine from matched glomerular centroids (robust to differing tract lengths)
c_src <- t(sapply(src, function(n) colMeans(nat::xyzmatrix(n))))
c_tgt <- t(sapply(tgt, function(n) colMeans(nat::xyzmatrix(n))))
aff <- Morpho::computeTransform(c_tgt, c_src, type = "affine")
src_aff <- lapply(src, function(n) { nat::xyzmatrix(n) <- Morpho::applyTransform(nat::xyzmatrix(n), aff); n })
```

## 3. Multi-object Deformetrica L-R fit (tracts + landmark anchor)

One diffeomorphism over ALL matched tracts at once (`NonOrientedPolyLine` +
`Varifold`), regularised by a shared neuropil-centroid **landmark** object — the
`inc_tracts_lmarks` variant from deformetricaLR. Contrast with a
landmark-free fit (`inc_tracts_no_landmark`) by dropping the `landmarks` argument.

```{r fit}
fit <- deformetrica_register_multi(
  sources = src_aff, targets = tgt,
  kernel_width = 20,                       # microns; the deformation length-scale
  landmarks = list(source = c_src %*% t(aff[1:3, 1:3]) + matrix(aff[1:3, 4], nrow(c_src), 3, byrow = TRUE),
                   target = c_tgt),
  data_sigma = 0.5, max_iterations = 200, device = "cuda", verbose = TRUE)
```

## 4. Apply + GIF of the L-R flow

Warp the full flipped-right neuron set through the fitted L-R diffeomorphism, and
turn the geodesic-flow timepoints into a GIF.

```{r apply-gif}
right_LR <- lapply(src_aff, function(n)
  deformetrica_shoot(n, fit$control_points, fit$momenta, kernel_width = fit$kernel_width))

frames <- sort(Sys.glob(file.path(fit$output_dir, "*GeodesicFlow*tp_*.vtk")))
render <- function(vtk, png) {
  m <- read.vtk(vtk, item = "points")
  rgl::open3d(useNULL = TRUE)
  rgl::points3d(nat::xyzmatrix(do.call(c, tgt)), col = "grey70", size = 1)   # left targets
  rgl::points3d(m, col = "firebrick", size = 2)                              # flowing right
  rgl::snapshot3d(png, webshot = FALSE); rgl::close3d()
}
pngs <- vapply(seq_along(frames), function(i) { p <- sprintf("lr_%02d.png", i); render(frames[i], p); p }, "")
# gifski::gifski(pngs, "fafb_left_right.gif", delay = 0.15)
```

## 5. Validation — does the L-R warp improve cognate correspondence?

The registration is good if a flipped-and-warped right neuron matches its left
cognate better than the affine-only flip does. We score with NBLAST (higher =
better) over the held-out glomeruli, comparing affine-only vs warped.

```{r validation}
library(nat.nblast)
dp <- function(n) nat::dotprops(n / 1, k = 5)
mrr <- function(query_set, target_set) {
  s <- nat.nblast::nblast_allbyall(c(query_set, target_set))  # illustrative; use held-out split
  mean(diag(s[names(query_set), names(target_set)]))
}
affine_score <- mrr(lapply(src_aff, dp), lapply(tgt, dp))
warp_score   <- mrr(lapply(right_LR, dp), lapply(tgt, dp))
c(affine = affine_score, warp = warp_score)   # warp should exceed affine
```

**Acceptance criteria (iterate `kernel_width` / `data_sigma` against these):** the
warped cognate NBLAST beats the affine-only baseline on held-out glomeruli, and
L-R type consistency (assign each warped right neuron to its nearest left cognate)
rises. If a single whole-brain L-R warp cannot achieve this, register per-region
(e.g. antennal-lobe glomeruli separately), exactly the neuropil-pair fallback used
in the mosquito vignette.
