---
title: "Warping a mosquito brain onto the fly (JRC2018U) with Deformetrica"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Warping a mosquito brain onto the fly (JRC2018U) with Deformetrica}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
# The data-pull and Deformetrica chunks need the Insect Brain Database, a
# Deformetrica (>= 4.3) install and a GPU, so they are eval=FALSE by default.
# Run them locally after `find_deformetrica()` resolves.
```

## Goal

Register the whole-brain surface of the **mosquito _Aedes aegypti_** onto the
**_Drosophila_ JRC2018U** unisex template brain, as accurately as possible, and
render the geodesic flow of the transform as a GIF. Cross-species brain shapes
differ enormously, so we do this in two stages — an **affine initialisation**
(Morpho) to remove gross pose/scale differences, then a **non-rigid Deformetrica
diffeomorphism** (surface registration). We validate on **homologous neuropils**
(antennal lobe, mushroom body, central complex) that both brains share.

This mirrors the two-stage recipe in the package's `deform/` scripts: affine
first, deformable second.

```{r setup}
library(deformetricar)
library(nat)
library(nat.flybrains)   # JRC2018U template
library(Morpho)          # affine initialisation + ICP
library(Rvcg)            # mesh decimation / surface metrics
# install.packages("remotes"); remotes::install_github("natverse/insectbrainr")
library(insectbrainr)    # Aedes brain from the Insect Brain Database
```

## 1. The two brain surfaces

```{r meshes}
## Target: Drosophila JRC2018U whole-brain surface, as a mesh3d (in microns).
fly <- as.mesh3d(nat.flybrains::JRC2018U)

## Source: Aedes aegypti standard brain from the Insect Brain Database.
## insectbrainr exports (verified): insectbraindb_species_info,
## insectbraindb_read_brain, insectbraindb_read_neurons, insectbraindb_neuron_info.
## The brain comes as neuropil sub-surfaces; the union is the whole-brain mesh.
## NOTE: insectbrainr is older and its Insect Brain Database mesh queries may need
## a fix upstream (see the package dev notes) — if `insectbraindb_read_brain`
## errors, that is the PR to make.
aedes_species <- insectbraindb_species_info("Aedes aegypti")
aedes_brain   <- insectbraindb_read_brain(brain = aedes_species)   # hxsurf, neuropil sub-regions
aedes <- as.mesh3d(aedes_brain)                                    # whole-brain union
```

## 2. Affine initialisation (Morpho)

A diffeomorphism cannot (and should not) absorb a cross-species change of pose
and scale — that is the affine step's job. We drive it from **homologous neuropil
centroids** (robust landmarks present in both brains), then refine with a
surface ICP.

```{r affine}
homologous <- c("antennal_lobe", "mushroom_body", "central_body", "optic_lobe")
centroid <- function(brain, region) colMeans(nat::xyzmatrix(subset(brain, region)))
lm_moz <- t(sapply(homologous, centroid, brain = aedes_brain))
lm_fly <- t(sapply(homologous, centroid, brain = nat.flybrains::JRC2018U))

aff      <- Morpho::computeTransform(lm_fly, lm_moz, type = "affine")   # moz -> fly
aedes_aff <- Morpho::applyTransform(aedes, aff)
aedes_aff <- Morpho::icp(aedes_aff, fly, iterations = 30, type = "affine")  # surface refine
```

## 3. Non-rigid Deformetrica warp (surface registration)

Decimate both surfaces to a tractable, comparable vertex count, then fit an
unlabelled-surface diffeomorphism (`SurfaceMesh` + `Current` attachment — no
vertex correspondence required). `kernel_width` sets the deformation length-scale
(a few tens of microns for whole-brain shape; sweep it as a validation knob).

```{r warp}
src <- Rvcg::vcgQEdecim(aedes_aff, percent = 0.15)
tgt <- Rvcg::vcgQEdecim(fly,       percent = 0.15)

fit <- deformetrica_register(
  source = src, target = tgt,        # mesh3d in, SurfaceMesh/Current auto-selected
  kernel_width = 30, timepoints = 15, max_iterations = 200,
  device = "cuda", verbose = TRUE)
```

## 4. Apply the transform + GIF frames

`deformetrica_shoot()` flows any object through the fitted diffeomorphism and
returns the same class it was given — so we can warp the full-resolution mosquito
mesh (not just the decimated fit surface). The Deformetrica shooting output writes
one VTK per timepoint under `fit$output_dir`/`workdir`; those are the GIF frames.

```{r apply-gif}
aedes_warped <- deformetrica_shoot(aedes_aff, fit$control_points, fit$momenta,
                                   kernel_width = fit$kernel_width, device = "cuda")

## Render each geodesic timepoint to a frame, then assemble a GIF.
frames <- sort(Sys.glob(file.path(fit$output_dir, "*GeodesicFlow*tp_*.vtk")))
render_frame <- function(vtk, png) {
  m <- read.vtk(vtk, item = "points")
  rgl::open3d(useNULL = TRUE); rgl::points3d(m, size = 1)
  rgl::wire3d(fly, col = "grey80")            # fly target for reference
  rgl::snapshot3d(png, webshot = FALSE); rgl::close3d()
}
pngs <- vapply(seq_along(frames), function(i) { p <- sprintf("frame_%02d.png", i); render_frame(frames[i], p); p }, "")
# gifski::gifski(pngs, "mosquito_to_fly.gif", delay = 0.15)   # or magick::image_animate
```

## 5. Validation — homologous neuropils

The whole-brain surface distance alone is a weak criterion (a diffeomorphism can
over-fit the outer hull). The meaningful test is whether **homologous interior
neuropils land on their fly counterparts** after the warp. For each shared
neuropil we report the symmetric surface distance and a volumetric Dice overlap
between the warped mosquito neuropil and the fly neuropil.

```{r validation}
surface_dist <- function(a, b) mean(Rvcg::vcgClostKD(a, b)$quality)   # a->b nearest-vertex dist
dice <- function(a, b) {                                              # volumetric overlap
  ov <- Rvcg::vcgBool(a, b, "intersection")
  2 * Rvcg::vcgVolume(ov) / (Rvcg::vcgVolume(a) + Rvcg::vcgVolume(b))
}
score_neuropil <- function(region) {
  moz_w <- deformetrica_shoot(as.mesh3d(subset(aedes_brain, region)),
                              fit$control_points, fit$momenta, kernel_width = fit$kernel_width)
  fly_n <- as.mesh3d(subset(nat.flybrains::JRC2018U, region))
  data.frame(region, affine_dist = surface_dist(Morpho::applyTransform(as.mesh3d(subset(aedes_brain, region)), aff), fly_n),
             warp_dist = surface_dist(moz_w, fly_n), dice = dice(moz_w, fly_n))
}
val <- do.call(rbind, lapply(homologous, score_neuropil))
val   # warp_dist should beat affine_dist for each neuropil; dice should rise
```

**Acceptance criteria (iterate `kernel_width` / decimation against these):**

1. Every homologous neuropil's `warp_dist` < its `affine_dist` (the diffeomorphism
   improves on the affine init everywhere, not just the hull).
2. Mean neuropil Dice rises materially over the affine-only baseline.
3. The transform is a valid diffeomorphism (no folding) — check that the Jacobian
   of the flow stays positive (Deformetrica will not fold, but a too-small kernel
   with too-few iterations can leave the surfaces unmatched).

### Fallback — register comparable neuropils as query/target pairs

If the whole-brain warp cannot satisfy (1)–(2) (very different overall shape), drop
to **per-neuropil registration**: fit one diffeomorphism per homologous pair
(mosquito AL → fly AL, MB → MB, …) with the neuropil surfaces as the query/target,
which removes the competing whole-brain constraints and is the more honest test of
local homology.

```{r fallback}
fit_al <- deformetrica_register(
  source = as.mesh3d(subset(aedes_brain, "antennal_lobe")),
  target = as.mesh3d(subset(nat.flybrains::JRC2018U, "antennal_lobe")),
  kernel_width = 15, device = "cuda")
```
